---
title: "Re-Calculate to Events per Volume Unit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Re-Calculate to Events per Volume Unit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(flowdex)
src <- "https://github.com/bpollner/data/raw/main/bottled_water/bottled_water.zip"
td <- tempdir()
# this is downloading and unzipping the bottled water data if not already present:
flowdex::check_download_data(where = td, data_source = src, dsname = "bottled_water")
#
knitr::opts_knit$set(root.dir = paste0(tempdir(), "/bottled_water"))
```

If the FCM-machine used for data acquisition does have a volumetric measurement module and the fcs files contain volumetric data, it is possible to **re-calculate any extracted fluorescence distribution to events per volume unit**.

In other words, the y-axis of any fluorescence distribution can be displayed in a fixed scale, i.e. events per volume unit. This enables the **comparison of numbers on a fixed scale** across samples, groups or experiments, **instead** of being confined to **compare percentages** of sub-groups of an overall population.

## Set up Example

To fully appreciate the benefit of being able to re-calculate any fluorescence distribution to events per volume unit, let´s look at a practical example.

```{r}
library(flowdex)
src <- "https://github.com/bpollner/data/raw/main/bottled_water/bottled_water.zip"
td <- tempdir()
# this is downloading and unzipping the bottled water data if not already present:
flowdex::check_download_data(where = td, data_source = src, dsname = "bottled_Water")
#
setwd(paste0(td, "/bottled_water")) # set the working directory to the unzipped folder
```

The unzipped folder 'bottled_water' in tempdir() contains the standard folder structure required by `flowdex`.

#### Description of Dataset

-   All samples are water samples stained with cybergreen following the method as described in:\
    Prest, E. I., Hammes, F., Kotzsch, S., van Loosdrecht, M. C., & Vrouwenvelder, J. S. (2013). Monitoring microbiological changes in drinking water systems using a fast and reproducible flow cytometric method. Water Research, 47(19), 7131-7142 <https://doi.org/10.1016/j.watres.2013.07.051>

-   Three in Austria commercially available bottled waters were examined:\
    Alpquell, Evian, and Volvic.

-   It was the purpose of the experiment to compare the number of autochthonous bacteria in the water types and to distinguish the water types by their fluorescence profile.

-   From each water type, 9 samples were taken:

    -   For 3 samples, labelled 'GrA' (Group A), data acquisition was done for \~30 seconds.

    -   For 3 samples, labelled 'GrB' (Group B), data acquisition was done for \~60 seconds.

    -   For 3 samples, labelled 'GrC' (Group C), data acquisition was done for \~90 seconds.

-   A sample-ID string has been provided at the time of data acquisition, providing the class-variables

    -   C_waterName: The name of the bottled water

    -   C_Repl: The replicate, ranging from r1 to r3

    -   C_group: The group.

## Visualize Fluorescence Distributions

The main difference between samples within each water type in the dataset downloaded above is of course the different **time** data acquisition was done for --- 30, 60 or 90 seconds. Hence, for each group within a water type, considerable differences in the number of recorded events can be expected, resulting in different amplitudes of the fluorescence distributions when using the raw-events, i.e. **not** re-calculated to events per volume unit.

But as each of the 9 samples within each water type comes from the ***same*** **water**, their fluorescence distributions should be quite similar. This can be achieved by **re-calculating** the fluorescence distribution to sample volume, thereby equalizing the differences in time data acquisition was done for.

### Make Gating Set

The differences in recorded events can easily be visualized by using `makefdmat()` or `flowdexit()` and by setting the argument `rcv` ('**r**ecalculate **t**o **v**olume') to `FALSE`.

First, set up some custom color and custom linetype to optimize the graphical display:

```{r}
nrSa <- 9 # the number of samples per water type
nrSaGr <- 3 # the number of samples in each group (different lengths)
colp <- RColorBrewer::brewer.pal(3, "Dark2") # obtain 3 colors
# colors
cc <- c(rep(colp[1],nrSa), rep(colp[2],nrSa), rep(colp[3],nrSa))
col_aq <- colp[1] # the color for the 'Alpquell' water
col_ev <- colp[2] # the color for the 'Evian' water
col_vo <- colp[3] # the color for the 'Volvic' water
# linetype
silt <- c(rep(3,nrSaGr), rep(2,nrSaGr), rep(1,nrSaGr)) # custom linetype for a single water type
clt <- rep(silt, 3) # custom linetype for the complete dataset
# bandpass
bp <- c(1400, 2300) # the bandpass that we will user later on
```

Now use `makeAddGatingSet()` to produce a gating set from the fcs files:

```{r}
gs <- makeAddGatingSet()
```

You can look at the gated data by calling:

``` r
plotgates(gs, spl="C_waterName") # is exporting the pdf to '/plots'
```

### NOT Re-Calculating to Sample Volume {#not-recalc}

First visualize the raw-events, i.e. the fluorescence distribution **not** re-calculated to events per volume (achieved by setting `rcv` to FALSE):

```{r}
fdmat_no_calc <- applyBandpass(makefdmat(gs, rcv = FALSE), bp)
ti <- "Bottled Water, Raw Events"
plotFlscDist(fdmat_no_calc, ccol=cc, clt = clt, ti = ti, fns = "_raw", leg=FALSE, toPdf = FALSE) # we choose to NOT have a legend here 
```

To make the three sub-groups within each water type even better visible, we will plot each water type individually:

```{r}
fdmat_aq <- flowdexit(patt="Aq", rcv = FALSE, expo = FALSE, stf = FALSE)
ti <- "Bottled Water: Alpquell, Raw Events"
plotFlscDist(applyBandpass(fdmat_aq, bp), ccol=col_aq, clt=silt, ti=ti, toPdf = FALSE)
#
fdmat_ev <- flowdexit(patt="Ev", rcv = FALSE, expo = FALSE, stf = FALSE)
ti <- "Bottled Water: Evian, Raw Events"
plotFlscDist(applyBandpass(fdmat_ev, bp), ccol=col_ev, clt=silt, ti=ti, toPdf = FALSE)
#
fdmat_vo <- flowdexit(patt="Vo", rcv = FALSE, expo = FALSE, stf = FALSE)
ti <- "Bottled Water: Volvic, Raw Events"
plotFlscDist(applyBandpass(fdmat_vo, bp), ccol=col_vo, clt=silt, ti=ti, toPdf = FALSE)
```

This clearly shows that within each water type there are three distinct groups, caused of course by the different time data acquisition was done for in each group, resulting in a considerably different number of events recorded in each group.

### Re-Calculating to Sample Volume

As all samples within each water type should be *quite similar*, we can equalize the differences shown above by **re-calculating** all samples **to** their **sample volume**, i.e. the volume of sample that was acquired during data acquisition:

```{r}
fdmat_calc <- applyBandpass(makefdmat(gs), bp)
ti <- "Bottled Water, Re-calculated to Events / ml"
plotFlscDist(fdmat_calc, ccol=cc, clt = clt, ti = ti, leg = FALSE, toPdf = FALSE)
```

Even though within each water type the samples showed considerable differences in the number of recorded events ([see above](#not-recalc)), when re-calculating the events and the resulting fluorescence distributions to the volume of the sample that was acquired during data acquisition, we can obtain very similar fluorescence distributions for **each sample** within each water type.
