---
title: "Get Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Installation

Install release version from CRAN via

``` r
install.packages("flowdex")
```

Or download from github:

``` r
library(devtools) 
install_github(repo="bpollner/flowdex", ref="main")
```

## Initialize Settings File System {#initialize-settings-file-system}

Package `flowdex` makes use of the dynamic settings file system provided by package [`uniset`](https://bpollner.github.io/uniset/). This has to be initialised **just once** by calling

```{r}
library(flowdex)
flowdex::setup_settings(path.expand("~"))
```

Follow the on-screen instructions and **possibly re-start R**.\
From now on, various global settings and default values used by `flowdex` can conveniently be viewed and changed in the file `flowdex_settings.R` residing at

```{r}
path.expand("~")
```

## Download Tutorial Dataset {#download-tutorial-dataset}

In order to walk the user over the main-features of package `flowdex`, a small tutorial-dataset has been created, along with gating strategy files and the required polygon-gate definitions.\
Also, the workflow how to create the gating strategy file and the polygon gate definitions will be explained.

First, download the tutorial data from its Github repository:

```{r}
library(flowdex)
src <- "https://github.com/bpollner/data/raw/main/flowdex_tutorial/flowdex_tutorial.zip"
td <- tempdir()
# this is downloading and unzipping the tutorial data if not already present:
flowdex::check_download_data(where = td, data_source = src, dsname = "flowdex_tutorial")
```

#### Note: Description of Tutorial Dataset

-   All samples are tap water samples stained with cybergreen following the method as described in:\
    Prest, E. I., Hammes, F., Kotzsch, S., van Loosdrecht, M. C., & Vrouwenvelder, J. S. (2013). Monitoring microbiological changes in drinking water systems using a fast and reproducible flow cytometric method. Water Research, 47(19), 7131-7142 <https://doi.org/10.1016/j.watres.2013.07.051>
-   It was the purpose of the experiment to monitor the number of autochthonous bacteria in the water samples and their fluorescence distribution (LNA vs. HNA bacteria) over time.
-   A sample-ID string has been provided at the time of data acquisition, providing class and numerical variables describing each sample as described below.
-   There are two groups, experiment and control. The experiment group is denoted as 'GPos', the control group is denoted as 'GNeg'.
-   From each group, 24 samples were taken each day, from which 18 samples are included In this dataset.
-   fcs files from day 4, day 5 and day 6 are included (denoted as T4, T5, and T6).
-   The 18 samples per day from each group are further structured in 'thirds' and beaker number: th1, th2 and th3 for the first, second and third third of the 18 samples, b1, b2, ... up to b6 for the 6 samples (b for beaker) within each third.

------------------------------------------------------------------------

## Quickstart {#quickstart}

After having (only once) set up the [settings file system](#initialize-settings-file-system) and having downloaded the [tutorial dataset](#download-tutorial-dataset), you could dive straight in -- for a quickstart to immediately see what `flowdex` can do, call:

```{r}
library(flowdex)
setwd(paste0(td, "/flowdex_tutorial"))
fdmat <- flowdexit() # this might take a few seconds
fdmat@pData[1:5,] # to inspect volume and sample ID data
fdmat@cyTags[1:5,] # to inspect class- and numerical variables assigned to each sample
fdist <- fdmat[[1]] # to inspect fluorescence distribution
fdist[1:5, 130:134] # pick some random fluorescence intensities
```

Look at 'flowdexTutorial/rawdata/flscData_gateStrat.xlsx.xlsx' to see the fluorescence distribution exported to file, observe the various sheets there.

------------------------------------------------------------------------

## Prepare Folder Structure

Take it step by step, not interested in the quickstart above: We assume that data from a single experiment will be residing in a single folder -- the 'home' folder for this experiment.\
Within this home-folder, `flowdex` requires a simple folder structure to properly work. Create that folder structure now within a newly created experiment home-folder called 'tap_water_home':

```{r}
td <- tempdir()
exp_home <- paste0(td, "/tap_water_home") 
dir.create(exp_home)
flowdex::genfs(exp_home) # create the required folder structure in 'tap_water_home'
```

Continue to [data acquisition](https://bpollner.github.io/flowdex/articles/acquire_data.html), or straight to [Workflow 2](https://bpollner.github.io/flowdex/articles/workflow_2.html), where you can learn how to extract fluorescence distributions and how to visualize them.
